---
title: "Data manipulation"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
#load packages
library(learnr)
library(gradethis)
library(tidyverse)
library(knitr)
library(png)
library(grid)
library(magick)

#options
knitr::opts_chunk$set(echo = FALSE)

#code checking
gradethis_setup()

#load data used in exercises
baby_names <- c("Anne", "Fleur", "Johan", "Klaas", "Renee")
params <- c("Ka" = 0.5, "V1" = 15.0, "CL" = 2.0)
bootstrap <- c(0.49, 16.9, 3.5)
para_boot <- data.frame("parameters" = params, "bootstrap" = bootstrap)
data("Remifentanil", package = "nlme")
PK_data <- Remifentanil
```

```{r, include = F}
orig_opts_hook <- knitr::opts_hooks$get("tutorial")
opts_hook_override <- function(options) {
  options[["exercise"]] <- F
  orig_opts_hook(options)
}
knitr::opts_hooks$set(tutorial = opts_hook_override)
```

## Selection and subsetting

> **_NOTE:_**  This tutorial is still under construction. So far it only contains one way of subsetting data. The tutorial will be extended in the coming weeks. 

This tutorial covers the basics of data manipulation with base R and the packages dplyr and tidyr. We will first start with the topic of selection and subsetting data. 

### Dimensions

To select your data, you need to know in which dimension you are working. This is graphically shown in figure 1. If you are working with vectors, this is easy, as there is only one dimension, you will be selecting the different elements of the vector. With data frames you are working with two dimensions, which means you can either select either rows or columns, or both. With lists, it is even more complex, as you can be selecting within an element of the list, these elements can be vectors, data frames or even another list. 
```{r fig1, echo = FALSE, out.width = "100%", fig.asp = 0.5, fig.cap = "**Figure 1: Dimensions in data structures**"}
img <- readPNG("./Images/Dimensions.png")
grid.raster(img)
```

Please note that selecting and subsetting are closely related, however, they are **not** synonyms! If you use one of the methods, which are described in the following paragraphs, you can select data to change or delete, but you can also extract certain data, which is called subsetting. Look at the example below, where the first example is selecting, where data is selected and changed. The second example is subsetting two names, to be saved in another vector. However, the original vector is not changed. 

```{r code 0, echo = TRUE, eval = TRUE}
baby_names <- c("Anne", "Fleur", "Johan", "Klaas", "Renee")
print(baby_names)

#Select first element and change
baby_names[1] <- "Bas"
print(baby_names)

#Subset second and fifth element and save in another vector
girls <- baby_names[c(2,5)]
print(girls)
print(baby_names)
```

### Select using [] and index

The simplest way to select and subset is by using square brackets ([]). When you place these after an object name, you can select in different ways. Most important to remember is that the number of dimensions are of influence how you use these brackets! 

### Vector

Vector is the easiest, as it has only one dimension, so one number is required within the square brackets. A vector is created in the example below, which consist of 5 elements. The easiest way is to set a number between 1 and 5 in the square brackets, resembling the index of the element you want to extract. If you want multiple elements, you can indicate this with `c()`, or `:` if it is a sequence. If you want to remove an element, you can indicate this with a minus sign (-) in front of the element number. 

```{r code1, exercise = FALSE, eval=TRUE, echo = TRUE}
baby_names[1]
baby_names[c(1,5,3)]
baby_names[1:3]
baby_names[-3]
```

### Data frame

The same principle can also be applied on data frames. However, because you have two dimensions: rows and columns, you need to indicate to R in which dimension you are working. To separate the two dimensions, you use a comma, with the first number indicating the row and the second number indicating the column. If you leave a number out, it indicates to R that you are not selecting in that dimension and want to keep everything. A simple data frame is created below and printed to show you what it looks complete. 

```{r code2, exercise = FALSE, eval=TRUE, echo = TRUE}
baby_names <- c("Anne", "Fleur", "Johan", "Klaas", "Renee")
freq <- c(100, 88, 78, 95, 45)
baby_freq <- data.frame("names" = baby_names, "Frequency" = freq)
print(baby_freq)
```

Next, we are going to subset it. First select the first row, then the first column and next the first value: extracted from first row, first column. We see that the output also differs. When the first row is extracted, the data frame structure is kept. However, when we extract the first column or the first value, we see that a vector is returned. It is important to keep this in mind when you are working  with your own data, as already the simple subsetting methods differ between different data structures. Similarly to vectors, you can select multiple rows or columns by using `c()`.  

```{r code3, exercise = FALSE, eval=TRUE, echo = TRUE}
#first row
class(baby_freq[1, ])
#first column
baby_freq[ , 1]
#value from first row and first column
baby_freq[1, 1]

#Row 1,3 and 5, all columns
baby_freq[c(1,3,5),]
```

### List

A list is also one dimensional, and therefore only requires one number in the square brackets to select an element. However, if you want to select something from an element, you need to consider its dimensions. Look at the example below, where a list is created from the vector and data frame we used in the two previous examples. 

```{r code4, exercise = FALSE, eval=TRUE, echo = TRUE}
baby_names <- c("Anne", "Fleur", "Johan", "Klaas", "Renee")
freq <- c(100, 88, 78, 95, 45)
baby_freq <- data.frame("names" = baby_names, "Frequency" = freq)
baby_names_freq <- list("names" = baby_names, "baby_freq" = baby_freq)
print(baby_names_freq)
```

The vector with the baby names is the first element and the data frame with the names and frequencies is the second element of the list. If we want the fourth name from the vector *Klaas*, we have to first select the first element of the list and then the fourth element of the vector. However, keep attention to the output. When you use a single set of square brackets, you get a list as output, containing the element(s) you selected. When you use a double set of square brackets ([[]]), you will extract the element from the list. 

```{r code5, exercise = FALSE, eval=TRUE, echo = TRUE}
#first element of the list
baby_names_freq[1]

#fourth element of the vector
baby_names_freq[[1]][4]

```

### Test your knowledge

```{r quiz1, echo = FALSE}
quiz(caption = "Quiz - Selection and subsetting with index",
  question("Select the correct formula to select the third element from the vector *vect* with 10 elements",
    answer("vect[3,10]", message = "Two numbers indicate two dimensions, which is only the case with data frames"),
    answer("vect[10]", message = "We want the third element, not the tenth"),
    answer("vect[3]", correct = TRUE),
    answer("vect[[3]]", message = "You only use double square brackets with lists"),
    allow_retry = TRUE
  ),
  question("Select the correct formula to select the second, third and fourth row and the fifth and sixth column of the data frame *dat*",
    answer("dat[2:4, 5:6]", correct = TRUE),
    answer("dat[2:6]", message = "A data frame has two dimensions, so you should always include a comma to indicate in which dimension you are selecting"),
    answer("dat[[2:6]]", message = "You only use double square brackets with lists"),
    answer("dat[5:6, 2:4]", message = "The index/indices for the comma indicate the rows, after the comma indicate the columns. You have switched them around"),
    allow_retry = TRUE
  ),
  question("If I want to extract the second column of the data frame, which is the second element of my list *lis*, which formula do I use?",
    answer("lis[2][2]", message = "If you use single square brackets to select an element from the list, it will give the element in a list output. The next set of square brackets wil not work, because there is only one element in the list you have created."),
    answer("lis[[2]][2]", message = "You will extract the data frame with the first set of square brackets, but to select the second column in the data frame, you need to consider the two dimensions of the data frame"),
    answer("lis[[2]][ , 2]", correct = TRUE),
    answer("lis[2,2]", message = "A list only has one dimension, so you need to start with extracting the data frame from the list"),
    allow_retry = TRUE))
```


### Select using [] and/or names

Previously we looked at selecting and subsetting using the square brackets and the index. Now, we are going to take it one step further and use names to select. Names can be present in a vector, called a *named* vector, in data frames as row names and/or column names and in lists, as element names. Depending on the data structure, different methods can be used. 

### Data frame

In the example below, we see an data frame, which has in this case, both row and column names.

```{r code 6, eval = TRUE, echo = TRUE}
params <- c("Ka" = 0.5, "V1" = 15.0, "CL" = 2.0)
bootstrap <- c(0.49, 16.9, 3.5)
para_boot <- data.frame("parameters" = params, "bootstrap" = bootstrap)
print(para_boot)
```

With the names in data frames, there are two methods to select. The first method is to use the row and/or column names instead of the indices in the square brackets. The advantage of using names instead of indices, is that names don't change when you add or remove new rows/columns or reorder the rows/columns. 

```{r code 7, eval = TRUE, echo = TRUE}
#1st row
para_boot["Ka", ]

#2nd column
para_boot[ , "bootstrap"]

#value in 1st row, 1st column
para_boot["Ka", "parameters"]
```

The second method to select is using `$` followed by a column name. This method can only be used to select columns, not rows. This method is also unique for column names, as you cannot use indices after a dollar sign `$`. As explained before, you can use this way of selecting to select a column and replace its values, or subset a column and store it in another object. 

```{r code 8, eval = TRUE, echo = TRUE}
#1st column
para_boot$parameters

#2nd column
para_boot$bootstrap
```

### Creating new columns

Selecting columns of data frames is a commonly used aspect of your coding during data manipulation. You often base new columns on present columns in the data frame, think for example about calculating the BMI based on the weight and length column. If you calculate a new column based on another column, it will repeat the calculation for each row, based on the values in the other column from each row. Look at the example below. First you create a new column, and fill it with NA. Because you only provide one value, it will repeat itself until the whole column is filled. Next, you can define what should be in the new variable. In this case, the absolute difference between the parameter estimate and the bootstrap estimate. If you look at the last print of the data frame para_boot, you see that the difference column is calculated for each row. 

```{r code 9, eval = TRUE, echo = TRUE}
print(para_boot)

#first create a new column and fill it with NA
para_boot$difference <- NA
print(para_boot)

#fill the new column with a calculation based on other column(s)
para_boot$difference <- abs(para_boot$parameters - para_boot$bootstrap)
print(para_boot)
```

Try to do this yourself as well, with the following tips:

- Take a look at the data frame *PK_data*  
- Create a new column, called *BMI* and fill it with `NA`
- For each row, calculate the BMI based, on the height `Ht` and weight `Wt` columns
- CHECK the values of the `Ht` and `Wt` columns, are they in the correct unit for the BMI calculation?

```{r exercise1, exercise = TRUE} 

```

```{r exercise1-hint-1} 
#you can use:
print(PK_data)
```

```{r exercise1-hint-2} 
#create a new column with:
PK_data$BMI
```

```{r exercise1-hint-3} 
#the formula for BMI is: weight(kg)/(height(m)^2)
#base the BMI columns on:
PK_data$Ht
PK_data$Wt
```

```{r exercise1-solution, exercise.reveal_solution = TRUE} 
#Solution
print(PK_data)
PK_data$BMI <- NA
PK_data$BMI <- PK_data$Wt / ((PK_data$Ht/100)^2)
```

```{r exercise1-check} 
gradethis::grade_code()
```
