---
title: "Plot creation"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
#load packages
library(learnr)
library(knitr)
library(gradethis)
library(png)
library(grid)
library(tidyverse)
library(vdiffr)
#library(emo) # devtools::install_github("hadley/emo")
library(patchwork)
library(gridExtra)

#options
knitr::opts_chunk$set(echo = FALSE)

#code checking
gradethis_setup()

#load data used in exercises
example_data <- read.table(file = "./data/example_data.txt", header = TRUE, sep = ",")
example_data_ID3 <- example_data %>% 
  filter(ID == 3) 

test_plots <- function(x, y) {
  fx <- tempfile()
  fy <- tempfile()
  on.exit({
    unlink(c(fx, fy))
  })
  vdiffr::write_svg(x, fx)
  vdiffr::write_svg(y, fy)
  identical(readLines(fx), readLines(fy))
}

p_data <- ggplot(data = example_data, aes(x = time, y = concentration))

p_predictions <- ggplot(data = example_data, aes(x = prediction, y = concentration)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Predicted concentration", y = "Observed concentration") +
  theme_bw()
p_population <- ggplot(data = example_data, aes(x = sex, y = age, fill = sex)) +
  geom_violin() +
  scale_fill_manual(values = c("#D81B60", "#FFC107")) +
  labs(x = "Sex", y = "Age") +
  theme_bw()
p_PK <- ggplot(data = example_data, aes(x = time, group = ID, color = as.factor(ID))) +
  geom_line(aes( y = prediction)) +
  geom_point(aes(y = concentration), alpha = 0.3) +
  scale_y_log10() +
  labs(x = "Time", y = "Concentration", color = "ID") +
  theme_bw()

```

```{r, include = F}
# orig_opts_hook <- knitr::opts_hooks$get("tutorial")
# opts_hook_override <- function(options) {
#   options[["exercise"]] <- F
#   orig_opts_hook(options)
# }
# knitr::opts_hooks$set(tutorial = opts_hook_override)
```


## Introduction 

Visualization is one of the strongest tools for data exploration and analysis. R contains a versatile library for plotting your data: `ggplot2`. This tutorial will guide you through the basics of using `ggplot2`, so at the end only your own imagination will stand between you and your data visualization.

### Grammar of ggplot

ggplot has its own grammar, which falls within the tidyverse framework (along with `dplyr` and `tidyr`). It works with layers (Figure 1), where you keep adding functions to create the plot you want. Table 1 gives an overview of the different functions that are part of ggplot. In the next sections you will see how to use these building blocks in creating real plots.

```{r fig1, echo = FALSE, out.width = "100%", fig.asp = 0.6, fig.cap = "**Figure 1: ggplot layers**"}
img <- readPNG("./Images/ggplot_layers.png")
grid.raster(img)
```

layer            
-------                       ----------------------- 
`data`                        add a data frame with variables you want to plot
`aes`                         aesthetic, map the variables to axes: x, y, but also color, point size etc.
`geom_*()`                    the type of plot (geometric object): points, lines, boxplot, barplot, etc.
`facet_*()`                   splitting up of data over different panels (or facets)
`scale_*()` and `coord_*()`   adjust the scales of plot: limits, transformations, color scale etc.
`theme_*()`                   adjust visual aspect: font size, background color, etc.

Table: Table 1: different layers in ggplot. `*` can be replaced by the specific element you want to plot, e.g. `geom_point()` for a scatter plot

You can find many more layer options in the [ggplot2 cheat sheets](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

### Data used in this tutorial

All exercises can be done within the tutorial. If you, however, want to try it in your own Rstudio (or on the server), follow these steps:

Step 1: download the data file used for this tutorial. 

  + click on [this link](https://raw.githubusercontent.com/vanhasseltlab/tutoRials/main/4_R_plot_creation/data/example_data.txt),
  + click right mouse button and click 'Save as ...' OR press ctrl + s (on mac: cmd + s)
  + save the file to a place on your computer where you want to access it from (`path_to_file`)


If you work on the server you can upload the file to the server after downloading it to your own machine.

Step 2: use the following code to load the data into your environment. Replace `path_to_file` with your path to the file where you downloaded it.

`example_data <- read.table(file = "path_to_file/example_data.txt", header = TRUE, sep = ",")`


## Simple plot

Lets start with building our first plot! We will need some data to begin with. The data set called `example_data` was added to this tutorial. We can look at what it contains using `head()`.

```{r code1, echo = TRUE}
head(example_data)
```


Lets make our first simple plot, containing the concentration (`y`) over time (`x`). The main function is `ggplot()`, here you can specify the data (`data = example_data`) and the aesthetic (`aes(x = time, y = concentration)`). Let us see what happens!

```{r code2, echo = TRUE}
ggplot(data = example_data, aes(x = time, y = concentration))
```

What a disappointment, you cannot see the data. For this, we first need to add a geometry layer.

### Geom(etry) layers

We can start adding layers using an addition sign `+`. Lets start with a point (or scatter) plot. 


```{r code3, echo = TRUE}
ggplot(data = example_data, aes(x = time, y = concentration)) +
  geom_point()
```

*the `aes()` can be specified either inside `ggplot()` or in a specific layer (e.g. `geom_point(aes(y = concentration))`).

### Plot objects

Plots from ggplot can be stored as an object. The object is not printed, but you can print it later on.

```{r code3a, echo = TRUE}
p_data <- ggplot(data = example_data, aes(x = time, y = concentration))
```

Once you have the ggplot object, you can add layers to it later.

```{r code3b, echo = TRUE, eval = TRUE}
p_data + geom_point()
```

### Exercise

Now we can start playing with the different layers! First we do some data manipulation using `dplyr`. The data contains multiple measurements per individual, so lets use the data for only individual `3`.


```{r code4, echo = TRUE}
example_data_ID3 <- example_data %>% 
  filter(ID == 3)

```

For this smaller data set, to visualize a model fit, we want to not only plot the observed `concentration` as points, but also the model `prediction` as **line** over time. 
- Add a line geometry to the plot below with on the `y` axis the `prediction`.

```{r exercise1, exercise = TRUE}
ggplot(data = example_data_ID3, aes(x = time)) +
  geom_point(aes(y = concentration))
```


```{r exercise1-check}
grade_result(
  pass_if(~ test_plots(.result, ggplot(data = example_data_ID3, aes(x = time)) +
                                  geom_point(aes(y = concentration)) +
                                  geom_line(aes(y = prediction))), 
          "You did it!"),
  pass_if(~ test_plots(.result, ggplot(data = example_data_ID3, aes(x = time)) +
                                  geom_line(aes(y = prediction)) +
                                  geom_point(aes(y = concentration))), 
          "You did it!"),
  fail_if(~ TRUE, 
          "Not yet there, hint: check out the geom_line() function help page.")
)
```


## Complex plot

The plots shown before can (more) easily be made using the base R plot function. But now we will truly explore the possibilities of ggplot.

### Grouping on variables

In concentration time plots for multiple individuals, you might like to add lines for the predictions. So lets try it!

```{r code5a, echo = TRUE}
ggplot(data = example_data, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction))
```

That looks very weird, instead of adding a line for each individual, ggplot adds a line over all individuals. Luckily ggplot has some very smart option for this: you can group geometries based on variables.

```{r code5b, echo = TRUE}
ggplot(data = example_data, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID))
```

Another benefit of grouping is that you can also group by two variables, with the interaction option. This can for instance be useful to see differences between two hospitals, which have used the same ID for their patients. We see more lines, because now we have a seperate line for each ID at each hospital. This can be made a lot clearer by using colors.

```{r code5c, echo = TRUE}
example_data_interaction <- example_data 
example_data_interaction$site <- NA
example_data_interaction$site[] <- c("Hospital A", "Hospital B")

ggplot(data = example_data_interaction, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = interaction(ID, site)))
```

### Add colors to your plot

You can also use color on a layer to let it pop.

```{r code6, echo = TRUE}
ggplot(data = example_data, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID), color = "red")
```

You can use colors to distinguish between individuals, by adding the variable ID as an aesthetic to `aes()`. **Be aware!** There is a difference here between placing this feature inside or outside of `aes()`!

```{r code7a, echo = TRUE}
ggplot(data = example_data, aes(x = time, color = ID)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID))
```

The above plot looks a bit weird. ggplot sees the ID variable as numeric. Try changing ID into a factor (`as.factor(ID)`).

```{r code7b, echo = TRUE}
ggplot(data = example_data, aes(x = time, color = as.factor(ID))) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID))
```



### Multiple plots: facetting

Lastly, you can facet based on different variables. Facets are different panels, each containing their own part of the data.You use the `~` operator, which means 'by' here. So if we want to group by individual: we add `~ ID` to the facet layer. Note that you don't need an `aes()` function inside the facet, it works separately from the geometries.

```{r code8, echo = TRUE}
ggplot(data = example_data, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID), color = "red") +
  facet_wrap(~ ID)
```

Please do not hesitate to play around with these different features before taking the quiz:

```{r exercise3a, exercise = TRUE}
ggplot(data = example_data, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID), color = "red") +
  facet_wrap(~ ID)
```



### Quiz

- See the figure below. They all have a `color` argument in the creation of the plot. For every panel: is the color given inside or outside of `aes()`?

```{r fig2, echo = FALSE, out.width = "100%", fig.asp = 0.6}
img <- readPNG("./Images/quiz_complex_plots.png")
grid.raster(img)
```


```{r quiz1, echo = FALSE}
quiz(caption = "", 
  question("Plot A:",
    answer("in `aes()`"),
    answer("outside `aes()`", correct = TRUE),
    incorrect = "Wrong!" 
  ),
  question("Plot B:",
    answer("in `aes()`", correct = TRUE),
    answer("outside `aes()`"),
    incorrect = "Wrong!" 
  ),
  question("Plot C:",
    answer("in `aes()`", correct = TRUE),
    answer("outside `aes()`"),
    incorrect = "Wrong!" 
  ),
  question("Plot D:",
    answer("in `aes()`", correct = TRUE),
    answer("outside `aes()`"),
    incorrect = "Wrong!" 
  ),
  question("Plot E:",
    answer("in `aes()`"),
    answer("outside `aes()`", correct = TRUE),
    incorrect = "Wrong!" 
  ),
  question("Plot F:",
    answer("in `aes()`", correct = TRUE),
    answer("outside `aes()`"),
    incorrect = "Wrong!" 
  )
)
```



### Exercises

#### Exercise 1
- Replace question marks (`'?'`) in the code to recreate the plot picture of the plot below.

```{r exercise2, exercise = TRUE, eval = FALSE, warning = FALSE}
ggplot(data = example_data, aes(x = time, y = prediction)) +
  geom_line(aes(group = '?', color = '?')) +
  geom_smooth(aes(color = '?'))
```

```{r exercise2-hint-1}
# Look at how many lines are drawn and look back at the theory for grouping variables
# Is one line drawn through all points or are lines grouped on a certain variable? 
```

```{r exercise2-hint-2}
# The color argument within aes usually results in a legend, to show which colors describe which variable
```

```{r exercise2-hint-3}
# The variables you need to fill in are ID and sex, try to find out yourself which question mark(s) they replace
```

```{r exercise2-check}
ggplot(data = example_data, aes(x = time, y = prediction)) +
  geom_line(aes(group = '?', color = '?')) +
  geom_smooth(aes(color = '?'))

grade_result(
  pass_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  geom_line(aes(group = ID, color = sex)) +
                                  geom_smooth(aes(color = sex))), 
          "You did it!"),
  
  fail_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  geom_line(aes(group = 'ID', color = 'sex')) +
                                  geom_smooth(aes(color = 'sex'))), 
          "Not yet there, hint: remove the quotation marks (') around your variable names."),
  fail_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  geom_line(aes(group = '?', color = '?')) +
                                  geom_smooth(aes(color = '?'))), 
          "You did not get anywhere... Replace the '?'s in the above code"),
  
  fail_if(~ TRUE, 
          "Wrong!")
)
```

```{r fig3, echo = FALSE, out.width = "100%", fig.asp = 0.6}
img <- readPNG("./Images/result_exercise_complex_plots.png")
grid.raster(img)
```

#### Exercise 2

- Use `facet_wrap()` in the code below to create individual plots for separate `weight_group`s


```{r exercise3, exercise = TRUE, eval = FALSE, warning = FALSE}
ggplot(data = example_data, aes(x = time, y = prediction)) +
  geom_line(aes(group = ID))
```

```{r exercise3-check}
grade_result(
  pass_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  geom_line(aes(group = ID)) +
                                  facet_wrap(~ weight_group)), 
          "You did it!"),
  
  fail_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  facet_wrap(~ weight_group)), 
          "Oh no, don't remove the geometry!"),
  fail_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  geom_line(aes(group = ID)) +
                                  facet_wrap(~ ID)), 
          "Also looks nice, but wrong, you are supposed to facet on weight_group"),
  fail_if(~ test_plots(.result, ggplot(data = example_data, aes(x = time, y = prediction)) +
                                  geom_line(aes(group = ID))), 
          "You did not get anywhere... Try facetting on weight_group"),
  
  fail_if(~ TRUE, 
          "Wrong!")
)
```



## Beautiful plot

You have probably been wondering up till this point: why is there such an ugly gray background in all of our plots? Luckily ggplot has a lot of options for adjusting the visuals of your plots, not only the color of the lines, but in labels, aspect ratio, background, sizes of different elements, etc.

### Aesthetics labels 

You can adjust the labels for the x-axis, y-axis, give a title, even change the title  all through one simple layer: `labs()`.

```{r code9, echo = TRUE}
ggplot(data = example_data, aes(x = time, color = sex)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID)) +
  labs(title = "PK example data plot", y = "Concentration (mg/l)",
       x = "Time (days)", color = "Sex")
```


### Scaling

You can scale your aesthetics, both the limits of your x and y axis and the colors you want your plot to have, through `scale_*()` functions. Let us try to zoom in on only the first 50 days of the data, by scaling x. The warnings show that ggplot actually removes all data points beyond time point 50. 

```{r code10, echo = TRUE}
ggplot(data = example_data, aes(x = time)) + 
  geom_point(aes(y = concentration)) +
  geom_line(aes(y = prediction, group = ID)) +
  scale_x_continuous(limits = c(0, 50))
```

We have already seen you can choose a color outside of `aes()` (e.g. using 'red', 'blue', '#D81B60'), but it is also possible to choose colors for `aes()`, using scaling.

```{r code11, echo = TRUE}
ggplot(data = example_data, aes(x = time)) +
  geom_line(aes(y = prediction, group = ID), color = "grey65") +
  geom_point(aes(y = concentration, color = sex)) +
  scale_color_manual(values = c("#D81B60", "#FFC107"))
```

The `scale_*()` functions contain two elements: the aesthetic you want to affect (e.g. `color` or `x`), and some last part, what type of scaling you are aiming for (e.g. `manual` or `continuous`). The last part depends on what the layer is like (is x continuous or discrete? Do you want to manually choose colors, or use some gradient?). You can also use transformations such as `scale_y_log10()` to log scale your plot.

There are many different scaling types, so for now I can advice you to check out the [ggplot2 cheat sheets](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) or help pages/internet sources.

### Themes

Now, lets finally get rid of the background. Themes can be used for many other functions, like font sizes, positioning of the legend, etc. Almost always when you want to achieve something visual, you will see forum posts with recommendations to adjust something in `theme()`. Here you see an example:

```{r code12, echo = TRUE}
ggplot(data = example_data, aes(x = time)) +
  geom_line(aes(y = prediction, group = ID), color = "grey65") +
  geom_point(aes(y = concentration, color = sex)) +
  scale_color_manual(values = c("#D81B60", "#FFC107")) +
  theme(legend.position = "top", 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "black"))
```

The options used above are: 

- `legend.position = "top"`, moving the legend to the top of the plot
- `panel.grid.minor = element_blank()`, removing the minor grid lines (or set to an empty/blank element)
- `panel.background = element_rect(fill = "black")`, changing the background color to black

You can play around with these if you want

```{r code13, exercise = TRUE}
ggplot(data = example_data, aes(x = time)) +
  geom_line(aes(y = prediction, group = ID), color = "grey65") +
  geom_point(aes(y = concentration, color = sex)) +
  scale_color_manual(values = c("#D81B60", "#FFC107")) +
  theme(legend.position = "top", 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "black"))
```

There is way more to explore here, but there are luckily some preset themes that you can use to easily change multiple plot aspects. A favorite of mine is `theme_bw()`, which gives you a clean white background and also changes the legends (white background) and grid (white to light gray).

```{r code14, echo = TRUE}
ggplot(data = example_data, aes(x = time)) +
  geom_line(aes(y = prediction, group = ID), color = "grey65") +
  geom_point(aes(y = concentration, color = sex)) +
  scale_color_manual(values = c("#D81B60", "#FFC107")) +
  theme_bw()
```


There are many more [predefined themes such as theme_void(), theme_minimal(), theme_classic(), etc](https://ggplot2.tidyverse.org/reference/ggtheme.html). 

And please, when giving presentations or writing papers: at least add `theme_bw()` to your plot to avoid those ugly gray backgrounds!

### Exercise

#### Exercise 1

- Let us shush up a plot. Try to recreate the plot below! The geometries and aesthetics are already present. Try building on this using labs, scale, and (predefined) themes.

```{r exercise4, exercise = TRUE, eval = FALSE, warning = FALSE, message = FALSE}
ggplot(data = example_data, aes(x = prediction, y = concentration)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red")
```

```{r exercise4-hint-1} 
# Four lines of code need to be added to recreate this figure
# Look at all the differences you see when you run the given code vs the given plot below
```

```{r exercise4-hint-2} 
# Look at the labels on the axes, the scale of the axes and the background
```

```{r exercise4-hint-3} 
# Background stuff is usually set by with the `theme()` function
# Apply the `theme_bw()` function you have read about in the theory
```

```{r exercise4-hint-4} 
# The lines in the background is called the grid, these options can be set in another `theme()` function
theme(panel.grid.minor = ...)
# Google and fill in on the ellipsis (...) what you need so it removes these minor grid lines
```

```{r fig4, echo = FALSE, out.width = "100%", fig.asp = 0.6}
img <- readPNG("./Images/exercise_beautiful_plots.png")
grid.raster(img)
```

```{r exercise4-check}
grade_result(
  pass_if(~ test_plots(.result, ggplot(data = example_data, aes(x = prediction, y = concentration)) +
                                  geom_point() +
                                  geom_abline(intercept = 0, slope = 1, color = "red") +
                                  labs(x = "Predicted concentration", y = "Observed concentration") +
                                  scale_x_continuous(limits = c(0, 50)) +
                                  theme_bw() +
                                  theme(panel.grid.minor = element_blank())), 
          "You did it!"),
  fail_if(~ test_plots(.result, ggplot(data = example_data, aes(x = prediction, y = concentration)) +
                                  geom_point() +
                                  geom_abline(intercept = 0, slope = 1, color = "red")), 
          "Come on, try adding some extra layers!"),
  fail_if(~ TRUE, 
          "Not there yet...")
)
```


#### Exercise 2

- Build your most beautiful creation!

```{r exercise5, exercise = TRUE, eval = FALSE, warning = FALSE}
ggplot(data = example_data)
```

```{r exercise5-check}
grade_result(
  pass_if(~ any(class(.result) %in% c("gg", "ggplot")) & !test_plots(.result, ggplot(data = example_data)), 
          "Very... creative!"),
  fail_if(~ test_plots(.result, ggplot(data = example_data)), 
          "Come on, try adding some aesthetics and geometries!"),
  fail_if(~ TRUE, 
          "This is not a ggplot!")
)
```



## Combining plots

Sometimes you want to show multiple plots together in one figure (for example when the journal you want to submit to has a figure limit).

For example, we want to show some diagnostics about the study population, the model fit and the PK curve together. What we first want to do, is save every plot as an object.

```{r code15, echo = TRUE}
p_predictions <- ggplot(data = example_data, aes(x = prediction, y = concentration)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Predicted concentration", y = "Observed concentration") +
  theme_bw()

p_population <- ggplot(data = example_data, aes(x = sex, y = age, fill = sex)) +
  geom_violin() +
  scale_fill_manual(values = c("#D81B60", "#FFC107")) +
  labs(x = "Sex", y = "Age") +
  theme_bw()

p_PK <- ggplot(data = example_data, aes(x = time, group = ID, color = as.factor(ID))) +
  geom_line(aes( y = prediction)) +
  geom_point(aes(y = concentration), alpha = 0.3) +
  scale_y_log10() +
  labs(x = "Time", y = "Concentration", color = "ID") +
  theme_bw()

```

You can now use print to get these figures printed.

```{r code16, exercise = TRUE}
print(p_predictions)
```

Now we want to stack these three together, maybe the PK on the left and the population and prediction plot stacked on the right. But how?

### Patchwork

There are two ways that I will show you here today. First using the package `patchwork`. You can install it directly. 

```{r code17, eval = FALSE, echo = TRUE}
install.packages("patchwork")
```

Patchworks uses very simple code to create different arrangements of your plots, using `+` to put plots next to each other, and `/` to put plots on top of each other.

```{r code18, echo = TRUE, eval = TRUE}
p_PK + (p_predictions / p_population)
```

Brackets can be used to give the order of operations.

```{r code19, echo = TRUE, eval = TRUE}
(p_PK + p_predictions) / p_population

```

Patchwork is very intuitive to use, so certainly check it out if you want to combine plots!

### gridExtra and `grid.arrange()`

The package `gridExtra` contains a function called `grid.arrange()`. There is a lot of smart code behind the ggplot package and grid.arrange works both with full ggplots and parts of ggplots, called grobs. This is way beyond the scope of this tutoRial, but when you get in to grobs, you know you are an expert.

Grid arrange uses a bit more complicated code and will line up the plots differently from patchwork. This is how you can use it:

```{r code20, echo = TRUE, eval = TRUE}
grid.arrange(p_PK, p_predictions, p_population, layout_matrix = rbind(c(1, 2), c(1, 3)))
```

You give the function the three plots as arguments and you add a matrix layout, Where the numbers correspond with the argument order, so $1$ = `p_PK`, $2$ = `p_predictions` and $3$ = `p_population`. In this case:

$$\begin{bmatrix}
1 & 2\\
1 & 3
\end{bmatrix}$$


Grid arrange is a nice way of making more complicate operations, such as adding labs on the fly, or changing the lay-out matrix:

$$\begin{bmatrix}
1 & 1 & 2\\
1 & 1 & 2\\
1 & 1 & 3
\end{bmatrix}$$



```{r code21, echo = TRUE, eval = TRUE}
grid.arrange(p_PK + labs(tag = "A"), 
             p_predictions + labs(tag = "B"), 
             p_population + labs(tag = "C"), 
             layout_matrix = rbind(c(1, 1, 2), c(1, 1, 2), c(1, 1, 3)))
```


Feel free to play around with these functions yourself!

```{r code22, exercise = TRUE}
grid.arrange(p_PK + labs(tag = "A"), 
             p_predictions + labs(tag = "B"), 
             p_population + labs(tag = "C"), 
             layout_matrix = rbind(c(1, 1, 2), c(1, 1, 2), c(1, 1, 3)))
```

### End of plots tutorial

Well done! You finished this tutorial about R! Please feel free to revisit parts that were difficult, or continue with the next tutorial "R: RxODE".

I would appreciate if you could [provide me feedback](https://forms.gle/2E7ifNqaPheYw7wU6) on this tutorial. This anonymously given feedback will be used to improve this tutorial.  

